Can Auxiliary Indicators Improve COVID-19 Forecasts and Hotspot Prediction?

### 2021-06-13 forecast/paper/paper.tex
Updates thru methods C (formerly D)

At the outset, we should be clear that our intent in this paper is \textit{not}
to provide an authoritative take on cutting-edge COVID-19 forecasting methods.
Instead, our purpose here is to provide a rigorous, quantitative assessment of
the utility that several auxiliary indicators---such as those derived from
internet surveys or medical insurance claims---provide in tasks that involve
predicting future trends in confirmed COVID-19 cases. To assess such utility in
as simple terms as possible, we center our study in the framework of a basic
autoregressive model (in which COVID cases in the near future are predicted from  
a linear combination of COVID cases in the near past), and ask whether the 
inclusion of an auxiliary indicator as an additional feature in such a model
improves predictive power. While forecasting carries a rich literature that
offers a wide range of techniques, see e.g., \cite{Hyndman:2018}, we purposely
constrain ourselves to very simple models, avoiding common enhancements such as
order selection, correction of outliers/anomalies in the data, and inclusion of
regularization or nonlinearities. That said, analyses of forecasts submitted to
the COVID-19 Forecast Hub \cite{ForecastHub} by a large community of modelers
have shown that simple, robust models have consistently been among the
best-performing over the pandemic \cite{Cramer:2021}, including time series
models similar to those we consider in what follows.

@@

We consider prediction of future COVID-19 case rates or case hotspots (to be
defined precisely shortly).  By case rate, we mean the case count per 100,000
people (the standard in epidemiology).  We use reported case data aggregated by 
JHU CSSE \cite{Dong:2020}, which is accessible through the COVIDcast API, like
the auxiliary indicators that we use to supplement the basic purely
autoregressive models.

The indicators we focus on provide information not generally available from
standard public health reporting. Among the many indicators in the API, we
consider the following five:
\begin{itemize}
\item Change Healthcare COVID-like illness (CHNG-CLI): The percentage of
  outpatient visits that are primarily about COVID-related symptoms, based on
  de-identified Change Healthcare claims data.
\item Change Healthcare COVID (CHNG-COVID): The percentage of outpatient visits
  with confirmed COVID-19, based on the same claims data.
\item COVID Tracking and Impact Survey CLI-in-community (CTIS-CLI-in-community): 
  The estimated percentage of the population who knows someone who is sick,
  based on Delphi's surveys of Facebook users.
\item Doctor Visits COVID-like illness (DV-CLI): The same as CHNG-CLI, but
  computed based on de-identified medical insurance claims from other health
  systems partners.  
\item Google Search Trends for anosmia and ageusia (Google-AA): A measure of 
  COVID-19 related Google search volume for queries related to anosmia (loss of
  smell), ageusia (loss of taste), or both.  
\end{itemize}
In short, we choose these indicators because, conceptually speaking, they
measure aspects of an individual's disease progression that would plausibly
precede the occurence of (at worst, co-occur  with) the report of a positive
COVID-19 test, through standard public health reporting streams.

@@

Vintage training data means different things, in practice, for different
signals. The three signals based on medical claims, CHNG-CLI, CHNG-COVID, and
DV-CLI, are most often 3-5 days latent, and subject to a considerable but
regular degree of revision or ``backfill'' after their initial publication date.
The survey-based signal, CTIS-CLI-in-community, is 2 days latent, and rarely
undergoes any revision at all.  The target variable itself, reported COVID-19
case rates, is 1 day latent, and exhibits frequent, unpredictable revisions
(sometimes clear anomalies) after initial publication.  Compared to the revision
pattern in the medical claims signals, which are appears to be much more
systematic in nature, the revision pattern in the case rates signal can be
highly erratic, resulting in big spikes as reporting backlogs are cleared,
changes in case definitions are made, other corrections are applied, etc.

TODO Are the above problems mentioned about case rates really to do with vintage 
data, or would they still exist in finalized data?

@@

\subsection{Missing Data Imputation}

@@

The first task we consider---\textit{forecasting}---is to predict the HRR-level
COVID-19 case incidence rates, averaged over some 7-day time period in the
future (the average number of new daily reported cases per 100,000 people, over
the 7-day period).  The second task---\textit{hotspot prediction}---is to
predict a binary variable defined in terms of relative changes of the
aforementioned target variable. Why define the target variable here via 7-day 
averaging? The short answer is robustness: averaging stabilizes the case time
series, and accounts for uninteresting artifacts like weekend-weekday
differences in the series.  Note that we can of course equivalently view this
(equivalent up to a constant factor) as predicting HRR-level COVID-19 case
incidence rates \textit{summed} over some 7-day period in the future (the total
number of new reported cases per 100,000 people, over the 7-day period), or
predicting a binary variable defined in terms of its relative changes.

To fix notation, let $Y_{\ell,t}$ denote the 7-day trailing average of COVID-19
case incidence rates in location (HRR) $\ell$ and at time (day) $t$.  In 
forecasting, a time $t$, we seek to predict $Y_{\ell,t+a}$ for each ``ahead''
value $a=7,\ldots,21$. Meanwhile, in hotspot prediction, we seek to predict
$Z_{\ell,t+a}$, a binary indicator defined in terms of the relative difference 
$(Y_{\ell,t+a}-Y_{\ell,t+a-7}) / Y_{\ell,t+a-7}$, again for each
$a=7,\ldots,21$.

For each prediction date $t$, we use a 21-day trailing window of data to train
our forecast or hotspot prediction models (so, e.g., the trained models will
differ from those at prediction date $t-1$).  This is done to account for
(potential) nonstationarity.  For simplicity, the forecasting and hotspot
prediction models are always fit to data across all HRRs (i.e., the coefficients
in the models do not account for location-specific effects).

In our analysis, we let the prediction date $t$ run over each day in between
early June and December 31, 2020.  The precise start date differs for
forecasting and hotspots prediction; for each task it was chosen to be the
earliest date at which the data needed to train all models was available, which
ends up being (per our setup, with 21 days of training data, and lagged values
of signals for features as we will detail shortly) June 6, 2020 for forecasting,
and June 13, 2020 for hotspot prediction. (The bottleneck here is the
CTIS-CLI-in-community signal, which does not exist before early April 2020, when
the survey was first launched). The end date was chosen with a consideration
again to align both tasks as best as possible, and because few hotspots exist
post December 31, 2020, due to the general and gradual decline of the pandemic
in 2021.

\subsubsection{Task 1: Forecasting}

Recall $Y_{t,\ell}$ denotes the 7-day trailing average of COVID-19 case
incidence rates in location $\ell$ and at time $t$.  Separately for each
$a=7,\ldots,28$, to predict $Y_{\ell,t+a}$ for ahead value $a$, we consider a
simple probabilistic forecasting model of the form:     
\begin{equation}
\label{eq:forecast_ar}
\Quantile_\tau(Y_{\ell,t+a} \,|\, Y_{\ell,s}, s \leq t)  
= \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y_{\ell,t-7j}.  
\end{equation}
This model uses current case rates, and the case rates 7 and 14 days ago, in
order to predict (the quantiles of) case rates in the future.  We consider a
total of 7 quantile levels (chosen in accordance with the county-level quantile
levels suggested by the COVID-19 Forecast Hub),      
\begin{equation}
\label{eq:quantile_levels}
\tau \in \{0.025, 0.1, 0.25, 0.5, 0.75, 0.9, 0.975 \}.
\end{equation}
To estimate the coefficients, we use \textit{quantile regression}
\cite{Koenker:1978, Koenker:2005, Koenker:2006} fit separately for each $\tau$,
using data from all 306 HRRs, and within each HRR, using the most recent 21 days
of training data.  To be clear, this gives 6,426 training samples for each
quantile regression problem.

In addition to the above ``pure'' autoregression model, we consider five
probabilistic forecasting models of the form: 
\begin{multline}
\label{eq:forecast_ar_x}
\Quantile_\tau(Y_{\ell,t+a} \,|\, Y_{\ell,s}, X_{\ell,s}, s \leq t)  
= \\ \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y_{\ell,t-7j} + 
\sum_{j=0}^2 \gamma^{a,\tau}_j X_{\ell,t-7j},
\end{multline}
where $X_{\ell,t}$ denotes any one of the five auxiliary indicators---CHNG-CLI, 
CHNG-COVID, CTIS-CLI-in-community, DV-CLI, or Google-AA---at location $\ell$ and
time $t$. Note that we apply the same lags (current value, along with the values
7 and 14 days ago) for the auxiliary signals as we do for case rates. Estimation
proceeds just as before: we use the same 7 quantile levels as in
\eqref{eq:quantile_levels}, and again use quantile regression separately for
each $\tau$, fit using data from all 306 HRRs and a trailing window of 21 days
of training data.

At prediction time, in order to deal with possible crossing violations (a
predicted quantile at level $\tau$ exceeds a predicted quantile at level $\tau'
> \tau$), we apply a simple post-hoc sorting.  See Figure \ref{fig:trajectory}
for example forecast.

@@

\subsubsection{Task 2: Hotspot Prediction}

Define the binary indicator:
$$
Z_{t,\ell} = \indicator{Y^\Delta_{\ell,t} \geq 0.25},
$$
where we use the notation \smash{$Y^\Delta_{\ell,t} = (Y_{\ell,t} - Y_{\ell,
    t-7})/(Y_{\ell,t-7})$}. In other words, $Z_{t,\ell}=1$ if the number of
newly reported cases over the past 7 days has increased by at least 25\%
compared to the preceding week.  When this occurs, we call location $\ell$ a
\textit{hotspot} at time $t$.

We treat hotspot prediction as a binary classification problem and use a setup 
altogether quite similar to the forecasting setup described previously.
Separately for each $a=7,\ldots,28$, to predict $Z_{\ell,t+a}$, we consider a
simple logistic model:
\begin{equation}
\label{eq:hotspot_ar}
\logit(Z_{\ell,t+a} \,|\, Y_{\ell,s}, s \leq t)  
= \alpha^{a,\tau} + \sum_{j=0}^2 \^{a,\tau}_j Y^\Delta_{\ell,t-7j}.  
\end{equation}
In addition to this ``pure'' autoregression model, we also consider five
logistic models of the form: 
\begin{multline}
\label{eq:hotspot_ar_x}
\logit(Z_{\ell,t+a} \,|\, Y_{\ell,s}, X_{\ell,s}, s \leq t)  
= \\ \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y_{\ell,t-7j} + 
\sum_{j=0}^2 \gamma^{a,\tau}_j X_{\ell,t-7j},
\end{multline}
where we use \smash{$X^\Delta_{\ell,t} = (X_{\ell,t} - X_{\ell,
    t-7})/(X_{\ell,t-7})$}, and again $X_{\ell,t}$ stands for any of the five  
auxiliary indicators at location $\ell$ and time $t$.  We fit all models
\eqref{eq:hotspot_ar}, \eqref{eq:hotspot_ar_x} using logistic regression, 
pooling all 306 HRRs and using a 21-day trailing window for the training data.

@@

>>>>

At the outset, we should be clear that our intent in this paper is \textit{not}
to provide an authoritative take on cutting-edge COVID-19 forecasting methods.
Instead, our purpose here is to provide a rigorous, quantitative assessment of
the utility that several auxiliary indicators---such as those derived from
internet surveys or medical insurance claims---provide in tasks that involve
predicting future trends in confirmed COVID-19 cases. To assess such utility in
as simple terms as possible, we center our study in the framework of a basic
autoregressive model (in which COVID cases in the near future are predicted from  
a linear combination of COVID cases in the near past), and ask whether the 
inclusion of an auxiliary indicator as an additional feature in such a model
improves its predictions.

While forecasting carries a rich literature that offers a wide range of
techniques, see e.g., \cite{Hyndman:2018}, we purposely constrain ourselves to
very simple models, avoiding common enhancements such as order selection,
correction of outliers/anomalies in the data, and inclusion of regularization or
nonlinearities. That said, analyses of forecasts submitted to the COVID-19
Forecast Hub \cite{ForecastHub} by a large community of modelers have shown that
simple, robust models have consistently been among the best-performing over the
pandemic \cite{Cramer:2021}, including time series models similar to those we
consider in what follows.

@@

We consider prediction of future COVID-19 case rates or case hotspots (to be
defined precisely shortly).  By case rate, we mean the case count per 100,000
people (the standard in epidemiology).  We use reported case data aggregated by 
JHU CSSE \cite{Dong:2020}, which is accessible through the COVIDcast API
\cite{CovidcastAPI}, like the auxiliary indicators that we use to supplement the
basic purely autoregressive models.

The indicators we focus on provide information not generally available from
standard public health reporting. Among the many auxiliary indicators collected
in the API, we study the following five:  
\begin{itemize}
\item Change Healthcare COVID-like illness (CHNG-CLI): The percentage of
  outpatient visits that are primarily about COVID-related symptoms, based on
  de-identified Change Healthcare claims data.
\item Change Healthcare COVID (CHNG-COVID): The percentage of outpatient visits
  with confirmed COVID-19, based on the same claims data.
\item COVID Tracking and Impact Survey CLI-in-community (CTIS-CLI-in-community): 
  The estimated percentage of the population who knows someone who is sick,
  based on Delphi's surveys of Facebook users.
\item Doctor Visits COVID-like illness (DV-CLI): The same as CHNG-CLI, but
  computed based on de-identified medical insurance claims from other health
  systems partners.  
\item Google Search Trends for anosmia and ageusia (Google-AA): A measure of 
  COVID-19 related Google search volume for queries related to anosmia (loss of
  smell), ageusia (loss of taste), or both.  
\end{itemize}
In short, we choose these indicators because, conceptually speaking, they
measure aspects of an individual's disease progression that would plausibly
precede the occurence of (at worst, co-occur  with) the report of a positive
COVID-19 test, through standard public health reporting streams.

@@

Vintage training data means different things, in practice, for different
signals. The three signals based on medical claims, CHNG-CLI, CHNG-COVID, and
DV-CLI, are most often 3-5 days latent, and subject to a considerable but
regular degree of revision or ``backfill'' after their initial publication date.
The survey-based signal, CTIS-CLI-in-community, is 2 days latent, and rarely
undergoes any revision at all.  The target variable itself, reported COVID-19
case rates, is 1 day latent, and exhibits frequent, unpredictable revisions
(sometimes clear anomalies) after initial publication.  Compared to the revision
pattern in the medical claims signals, which are appears to be much more
systematic in nature, the revision pattern in the case rates signal can be
highly erratic. Big spikes or other anomalies can occur in the data as reporting
backlogs are cleared, changes in case definitions are made, etc. Groups like JHU 
CSSE then work tirelessly to correct such anomalies (e.g., back-distribute a
huge spike when a reporting backlog is cleared), resulting in highly nontrivial
and unpredictable revisions.

@@

@@

\begin{table*}[t]
  \centering
  \caption{Summary of forecasting and hotspot prediction tasks considered in
    this paper.}
  \begin{tabular}{l p{2.75in} p{2.75in}}
    \toprule
    & \textbf{Forecasting} & \textbf{Hotspot prediction} \\
    \midrule
    Response variable & $Y_{\ell,t}$ (7-day trailing average of COVID-19 case
 incidence rates, per location $\ell$ and time $t$) & $Z_{\ell,t} =
\indicator{Y_{\ell,t} \geq 1.25 \cdot Y_{\ell, t-7}}$ (indicator that $Y_{\ell,t}$ 
grows by more than 25\% relative to the preceding week) \\
    Geographic resolution & Hospital referral region (HRR) & Hospital referral
region (HRR) \\ 
    Forecast period & June 6--December 31, 2020 & June 13--December 31, 2020 \\  
    Model type & Quantile regression & Logistic regression \\
    Evaluation metric & Weighted interval score (WIS) & Area under curve (AUC) \\
    \bottomrule
  \end{tabular}
\label{tab:analysis_tasks}
\end{table*}

To fix notation, let $Y_{\ell,t}$ denote the 7-day trailing average of COVID-19
case incidence rates in location (HRR) $\ell$ and at time (day) $t$.  To be
clear, this is the number of new daily reported cases per 100,000
people, averaged over the 7-day period $t-6, \ldots t$. The first task we
consider---\textit{forecasting}---is to predict $Y_{\ell,t+a}$ for each
``ahead'' value $a=7,\ldots,21$.  The second task---\textit{hotspot
  prediction}---is to predict a binary variable defined in terms of the relative
change of $Y_{\ell,t+a}$ (relative to its value one week prior,
$Y_{\ell,t+a-7}$), again for each $a=7,\ldots,21$.

Why do we define the response variables via 7-day averaging? The short answer 
is robustness: averaging stabilizes the case time series, and accounts for 
uninteresting artifacts like weekend-weekday differences in the series.  Note
that we can of course equivalently view this (equivalent up to a constant
factor) as predicting the HRR-level COVID-19 case incidence rate \textit{summed} 
over some 7-day period in the future, and predicting a binary variable derived 
from this.

In what follows, we cover more details on our two analysis
tasks. Table~\ref{tab:analysis_tasks} presents a summary.

\paragraph{Dynamic Re-Training}

For each prediction date $t$, we use a 21-day trailing window of data to train
our forecast or hotspot prediction models (so, e.g., the trained models will
differ from those at prediction date $t-1$).  This is done to account for
(potential) nonstationarity.  For simplicity, the forecasting and hotspot
prediction models are always trained on data across all HRRs (i.e., the
coefficients in the models do not account for location-specific effects).

\paragraph{Prediction Period}

In our analysis, we let the prediction date $t$ run over each day in between
early/mid June and December 31, 2020.  The precise start date differs for 
forecasting and hotspots prediction; for each task it was chosen to be the
earliest date at which the data needed to train all models was available, which
ends up being (per our setup, with 21 days of training data, and lagged values
of signals for features as we will detail shortly) June 6, 2020 for forecasting,
and June 13, 2020 for hotspot prediction. (The bottleneck here is the
CTIS-CLI-in-community signal, which does not exist before early April 2020, when
the survey was first launched). The end date was chosen again with a
consideration to align both tasks as best as possible, and because few hotspots
exist post December 31, 2020, due to the general and gradual decline of the
pandemic in 2021.

\paragraph{Forecasting Models}

Recall $Y_{t,\ell}$ denotes the 7-day trailing average of COVID-19 case
incidence rates in location $\ell$ and at time $t$.  Separately for each
$a=7,\ldots,28$, to predict $Y_{\ell,t+a}$ for ahead value $a$, we consider a
simple probabilistic forecasting model of the form:     
\begin{equation}
\label{eq:forecast_ar}
\Quantile_\tau(Y_{\ell,t+a} \,|\, Y_{\ell,s}, s \leq t)  
= \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y_{\ell,t-7j}.  
\end{equation}
This model uses current case rates, and the case rates 7 and 14 days ago, in
order to predict (the quantiles of) case rates in the future.  We consider a
total of 7 quantile levels (chosen in accordance with the county-level quantile
levels suggested by the COVID-19 Forecast Hub),      
\begin{equation}
\label{eq:quantile_levels}
\tau \in \{0.025, 0.1, 0.25, 0.5, 0.75, 0.9, 0.975 \}.
\end{equation}
We fit \eqref{eq:forecast_ar} using \textit{quantile regression}
\cite{Koenker:1978, Koenker:2005, Koenker:2006} separately for each $\tau$, 
using data from all 306 HRRs, and within each HRR, using the most recent 21 days
of training data.  This gives us 6,426 training samples for each quantile 
regression problem.

In addition to the above ``pure'' autoregression model, we consider five
probabilistic forecasting models of the form: 
\begin{multline}
\label{eq:forecast_ar_x}
\Quantile_\tau(Y_{\ell,t+a} \,|\, Y_{\ell,s}, X_{\ell,s}, s \leq t)  
= \\ \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y_{\ell,t-7j} + 
\sum_{j=0}^2 \gamma^{a,\tau}_j X_{\ell,t-7j},
\end{multline}
where $X_{\ell,t}$ denotes any one of the five auxiliary indicators---CHNG-CLI, 
CHNG-COVID, CTIS-CLI-in-community, DV-CLI, or Google-AA---at location $\ell$ and
time $t$. Note that we apply the same lags (current value, along with the values
7 and 14 days ago) for the auxiliary indicators as we do for the case
rates. Training then proceeds just as before: we use the same 7 quantile levels
in \eqref{eq:quantile_levels}, and fit quantile regression separately for each 
level $\tau$, using data from all 306 HRRs and a trailing window of 21 days of
training data.

At prediction time, in order to deal with possible crossing violations (a
predicted quantile at level $\tau$ exceeds a predicted quantile at level $\tau'
> \tau$), we apply a simple post-hoc sorting.  See Figure \ref{fig:trajectory}
for an example forecast.

@@

\paragraph{Hotspot Prediction Models}

Define the binary indicator:
$$
Z_{t,\ell} = \indicator{Y^\Delta_{\ell,t} \geq 0.25},
$$
where we use the notation \smash{$Y^\Delta_{\ell,t} = (Y_{\ell,t} - Y_{\ell,
    t-7})/(Y_{\ell,t-7})$}. In other words, $Z_{t,\ell}=1$ if the number of
newly reported cases over the past 7 days has increased by at least 25\%
compared to the preceding week.  When this occurs, we say location $\ell$ is a
\textit{hotspot} at time $t$.

We treat hotspot prediction as a binary classification problem and use a setup 
altogether quite similar to the forecasting setup described previously.
Separately for each $a=7,\ldots,28$, to predict $Z_{\ell,t+a}$, we consider a
simple logistic model:
\begin{equation}
\label{eq:hotspot_ar}
\logit(Z_{\ell,t+a} \,|\, Y_{\ell,s}, s \leq t)  
= \alpha^{a,\tau} + \sum_{j=0}^2 \^{a,\tau}_j Y^\Delta_{\ell,t-7j}.  
\end{equation}
In addition to this ``pure'' autoregression model, we also consider five
logistic models of the form: 
\begin{multline}
\label{eq:hotspot_ar_x}
\logit(Z_{\ell,t+a} \,|\, Y_{\ell,s}, X_{\ell,s}, s \leq t)  
= \\ \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y^\Delta_{\ell,t-7j} +  
\sum_{j=0}^2 \gamma^{a,\tau}_j X^\Delta_{\ell,t-7j},
\end{multline}
where we use \smash{$X^\Delta_{\ell,t} = (X_{\ell,t} - X_{\ell,
    t-7})/(X_{\ell,t-7})$}, and again $X_{\ell,t}$ stands for any of the five  
auxiliary indicators at location $\ell$ and time $t$.  We fit all models
\eqref{eq:hotspot_ar}, \eqref{eq:hotspot_ar_x} using logistic regression, 
pooling all 306 HRRs and using a 21-day trailing window for the training data.

@@

\subsection{Other Considerations}

\paragraph{Missing Data Imputation}

\paragraph{Spatial Heterogeneity}

\paragraph{Backfill, Nowcasting}
